Relationships in the extremes and their influence on competition and coexistence: Supporting information

### 2021-07-07 paper/SupMat.Rnw
further progress in methods and associated sup mat

@@

&= - \E[\ln(1-\delta+\delta \exp(\sigma(\tilde{b}_j^\shp - \tilde{b}_j)))],

@@

Using the above formulas, we observe that all the terms of the decomposition of $\olr_{i \bs i}$ depend on 
$\delta$, $\sigma$, and $\mu_i - \mu_j$, but not separately on $\mu_i$ or $\mu_j$. The terms of the decomposition
of $\olr_{j \bs i}$ depend on $\delta$ and $\sigma$, but do not depend on $\mu_i$ or $\mu_j$ at all.
As discussed in the main text and SI section \ref{SIsect:scaling_factors}, we consider two alternative
values for the scaling factors, $q_{ij}$, namely $q_{ij}=1$ and 
\begin{align}
q_{ij} &= \frac{\exp(\mu_i)}{(1-\delta) \exp(\mu_j)+\delta \exp(\mu_i)} \\
&= \frac{\exp(-\mu_i)\exp(\mu_i)}{\exp(-\mu_i)[(1-\delta) \exp(\mu_j)+\delta \exp(\mu_i)]} \\
&= \frac{1}{(1-\delta)\exp(-(\mu_i-\mu_j))+\delta}.
\end{align}
These also either do not depend on $\mu_i$ or $\mu_j$ at all or they depend on the difference 
$\mu_i-\mu_j$. Thus each of the terms of the decomposition (\ref{MT-eq:full_decomp}) from the main text
depends only on $\mu_i-\mu_j$ and not separately on $\mu_i$ or $\mu_j$, as stated in the Methods section of the main text.

@@

Next, proceed as follows to get estimates of the terms in the decomposition (\ref{eq:decomp_overall_rjbsi}). 
\begin{enumerate}
\item Use $\epsilon_i^0=0$, as in (\ref{eq:decomp:baseline_j}), and this is the exact value.
\item Estimate $\olep_j^E$, following (\ref{eq:decomp:contrib_of_var_E_j}), as
$\widehat{\olep_j^E}=\mean_k[\ln(1-\delta+\delta\exp(\sigma u^{(k)}-\sigma^2/2))]-\epsilon_j^0$, an estimate
with standard error $\se \left( \widehat{\olep_j^E} \right) = \sd_k [\ln(1-\delta+\delta\exp(\sigma u^{(k)}-\sigma^2/2))]/\sqrt{M}$.
\item Estimate $\olep_j^C$, following (\ref{eq:decomp:contrib_of_var_C_j}), as
$\widehat{\olep_j^C}=\mean_k[\ln(1-\delta+\delta\exp(-\sigma u^{(k)}+\sigma^2/2))]-\epsilon_j^0$, which has 
standard error $\se \left( \widehat{\olep_j^C} \right) = \sd_k [\ln(1-\delta+\delta\exp(-\sigma u^{(k)}+\sigma^2/2))]/\sqrt{M}$.
\item Note that $\tilde{b}_j$ and $\tilde{b}_j^\shp$ in (\ref{eq:decomp:j:EshpC}) are standard normally distributed and
independent, so $\tilde{b}_j^\shp - \tilde{b}_j$ is normally distributed with mean $0$ and variance $2$. So we can 
estimate $\olep_j^{(E \shp C)}$, following (\ref{eq:decomp:j:EshpC}), as 
$\widehat{\olep_j^{(E \shp C)}} = \mean_k [\ln(1-\delta+\delta\exp(\sigma\sqrt{2}u^{(k)}))]-\widehat{\olep_j^E}-\widehat{\olep_j^C}-\epsilon_j^0$.
The standard error of this expression is $\se \left(  \widehat{\olep_j^{(E \shp C)}} \right) = 
\sd_k[\ln(1-\delta+\delta\exp(\sigma\sqrt{2}u^{(k)}))-\ln(1-\delta+\delta\exp(u^{(k)}-\sigma^2/2))-
\ln(1-\delta+\delta\exp(-\sigma u^{(k)}+\sigma^2/2))]/\sqrt{M}$.
\item Estimate
%***DAN: DAN, proceed from here
\end{enumerate}

@@

%***DAN: This and probably all or most of the following sections will have to be edited, not least to make the
%notation consistent with the notation now used in the Theory section of the main text
%***DAN: At the end of Theory, I mention that in methods the tricks for computational efficiency are
%elaborated, but actually most of those will be in a sup mat section, the material for which will be 
%an edited version of a lot of what is in this section.

>>>>

&= 0 \label{eq:decomp:j:EnatC}

@@

&= - \E[\ln(1-\delta+\delta \exp(\sigma(\tilde{b}_j^\shp - \tilde{b}_j)))],\label{eq:decomp:j:EChard}

@@

Using the above formulas, we observe that all the terms of the decomposition of $\olr_{i \bs i}$ depend on 
$\delta$, $\sigma$, and $\mu_i - \mu_j$, but not separately on $\mu_i$ or $\mu_j$. The terms of the decomposition
of $\olr_{j \bs i}$ depend on $\delta$ and $\sigma$, but do not depend on $\mu_i$ or $\mu_j$ at all.
As discussed in the main text and SI section \ref{SIsect:scaling_factors}, we consider two alternative
values for the scaling factors, $q_{ij}$, namely $q_{ij}=1$ and 
\begin{align}
q_{ij} &= \frac{\exp(\mu_i)}{(1-\delta) \exp(\mu_j)+\delta \exp(\mu_i)} \\
&= \frac{\exp(-\mu_i)\exp(\mu_i)}{\exp(-\mu_i)[(1-\delta) \exp(\mu_j)+\delta \exp(\mu_i)]} \\
&= \frac{1}{(1-\delta)\exp(-(\mu_i-\mu_j))+\delta}.\label{eq:decomp:qnot1}
\end{align}
These also either do not depend on $\mu_i$ or $\mu_j$ at all or they depend on the difference 
$\mu_i-\mu_j$. Thus each of the terms of the decomposition (\ref{MT-eq:full_decomp}) from the main text
depends only on $\mu_i-\mu_j$ and not separately on $\mu_i$ or $\mu_j$, as stated in the Methods section of the main text.

@@

Next, proceed as follows to get estimates of the terms in the decomposition (\ref{eq:decomp_overall_rjbsi}). 
\begin{enumerate}
\item Use $\epsilon_i^0=0$, as in (\ref{eq:decomp:baseline_j}), and this is the exact value.
\item Estimate $\olep_j^E$, following (\ref{eq:decomp:contrib_of_var_E_j}), as
$\widehat{\olep_j^E}=\mean_k[\ln(1-\delta+\delta\exp(\sigma u^{(k)}-\sigma^2/2))]-\epsilon_j^0$, an estimate
with standard error $\se \left( \widehat{\olep_j^E} \right) = \sd_k [\ln(1-\delta+\delta\exp(\sigma u^{(k)}-\sigma^2/2))]/\sqrt{M}$.
\item Estimate $\olep_j^C$, following (\ref{eq:decomp:contrib_of_var_C_j}), as
$\widehat{\olep_j^C}=\mean_k[\ln(1-\delta+\delta\exp(-\sigma u^{(k)}+\sigma^2/2))]-\epsilon_j^0$, which has 
standard error $\se \left( \widehat{\olep_j^C} \right) = \sd_k [\ln(1-\delta+\delta\exp(-\sigma u^{(k)}+\sigma^2/2))]/\sqrt{M}$.
\item Note that $\tilde{b}_j$ and $\tilde{b}_j^\shp$ in (\ref{eq:decomp:j:EshpC}) are standard normally distributed and
independent, so $\tilde{b}_j^\shp - \tilde{b}_j$ is normally distributed with mean $0$ and variance $2$. So we can 
estimate $\olep_j^{(E \shp C)}$, following (\ref{eq:decomp:j:EshpC}), as 
$\widehat{\olep_j^{(E \shp C)}} = \mean_k [\ln(1-\delta+\delta\exp(\sigma\sqrt{2}u^{(k)}))]-\widehat{\olep_j^E}-\widehat{\olep_j^C}-\epsilon_j^0$.
The standard error of this expression is $\se \left(  \widehat{\olep_j^{(E \shp C)}} \right) = 
\sd_k[\ln(1-\delta+\delta\exp(\sigma\sqrt{2}u^{(k)}))-\ln(1-\delta+\delta\exp(u^{(k)}-\sigma^2/2))-
\ln(1-\delta+\delta\exp(-\sigma u^{(k)}+\sigma^2/2))]/\sqrt{M}$.
\item Estimate $\olep_j^{[EC]}$, following (\ref{eq:decomp:j:EChard}), as $\widehat{\olep_j^{[EC]}}=\mean_k [-\ln(1-\delta+\delta \exp(\sigma \sqrt{2} u^{(k)}))],$ which has $\se \left( \widehat{\olep_j^{[EC]}} \right) = \sd_k[-\ln(1-\delta+\delta \exp(\sigma \sqrt{2} u^{(k)}))]/\sqrt{M}$.
\item Use $\olep_j^{[E \nat C]}=0$, following (\ref{eq:decomp:j:EnatC}). This is exact.
\item Finally, use $\olr_{j \bs i} = 0$, again exact. 
\end{enumerate}
As for (\ref{eq:decomp_overall_IGR}), it is again easy to see that (\ref{eq:decomp_overall_rjbsi}) applies for the 
estimates as well as for the quantities being estimated.

Next, $q_{ij}$ can be taken (see SI section \ref{SIsect:scaling_factors}) to be, alternatively, $1$ or 
the quantity displayed in (\ref{eq:decomp:qnot1}). Hence the terms of the decomposition 
(\ref{eq:decomp_overall_delta}) can straightforwardly be etimated using our estimates elaborated 
above, as follows. There are two alternative versions of the estiamtors for this decomposition, corresponding 
to the alternative values for $q_{ij}$.
\item Compute $\Delta_i^0 = \epsilon_i^0 - q_{ij} \epsilon_j^0$, which is computed exactly, since 
the quantities described above for the components $\epsilon_i^0$, $\epsilon_j^0$ and $q_{ij}$ are exact, as opposed to being
estimates. Simplifying, since $\epsilon_j^0=0$, we have $\Delta_i^0 = \epsilon_i^0$.
\item Estimate $\Delta_i^E$ as $\widehat{\Delta_i^E}=\widehat{\olep_i^E}-q_{ij}\widehat{\olep_j^E}=
\mean_k[\ln(1-\delta+\delta\exp(\sigma u^{(k)}+\mu_i-\mu_j-\sigma^2/2))]-\epsilon_i^0-
q_{ij} \mean_k[\ln(1-\delta+\delta\exp(\sigma u^{(k)}-\sigma^2/2))],$ which has standard error 
$\se\left( \widehat{\Delta_i^E} \right)=\sd_k[\ln(1-\delta+\delta\exp(\sigma u^{(k)}+\mu_i-\mu_j-\sigma^2/2))-
q_{ij} \ln(1-\delta+\delta\exp(\sigma u^{(k)}-\sigma^2/2))]/\sqrt{M}.$
\item Estimate $\Delta_i^C$ as $\widehat{\Delta_i^C}=\widehat{\olep_i^C}-q_{ij}\widehat{\olep_j^C}=
\mean_k[\ln(1-\delta+\delta\exp(-\sigma u^{(k)}+\mu_i-\mu_j+\sigma^2/2))]-
\epsilon_i^0-q_{ij}\mean_k[\ln(1-\delta+\delta\exp(-\sigma u^{(k)}+\sigma^2/2))],$ which has standard error
$\se \left( \widehat{\Delta_i^C} \right)=\sd_k[\ln(1-\delta+\delta\exp(-\sigma u^{(k)}+\mu_i-\mu_j+\sigma^2/2))-
q_{ij}\ln(1-\delta+\delta\exp(-\sigma u^{(k)}+\sigma^2/2))]/\sqrt{M}$.
\item Estimate $\Delta_i^{(E \shp C)}$ as $\widehat{\Delta_i^{(E \shp C)}} = \widehat{\olep_i^{(E \shp C)}}-q_{ij} \widehat{\olep_j^{(E \shp C)}}=
\mean_k[\ln(1-\delta+\delta\exp(\sigma\sqrt{2}u^{(k)}+\mu_i-\mu_j))]-\widehat{\olep_i^E}-\widehat{\olep_i^C}-\epsilon_i^0-
q_{ij}[\mean_k [\ln(1-\delta+\delta\exp(\sigma\sqrt{2}u^{(k)}))]-\widehat{\olep_j^E}-\widehat{\olep_j^C}-\epsilon_j^0].$
This simplifies to $\mean_k[\ln(1-\delta+\delta\exp(\sigma\sqrt{2}u^{(k)}+\mu_i-\mu_j))]-
q_{ij}\mean_k[\ln(1-\delta+\delta\exp(\sigma\sqrt{2}u^{(k)}))]-\widehat{\Delta_i^E}-\widehat{\Delta_i^C}-\Delta_i^0$, 
which can be straightforwardly shown has standard error
$\se \left( \widehat{\Delta_i^{(E \shp C)}} \right) = 
\sd_k[\ln(1-\delta+\delta\exp(\sigma\sqrt{2}u^{(k)}+\mu_i-\mu_j))-
\ln(1-\delta+\delta\exp(\sigma u^{(k)}+\mu_i-\mu_j-\sigma^2/2))-
\ln(1-\delta+\delta\exp(-\sigma u^{(k)}+\mu_i-\mu_j+\sigma^2/2))-
q_{ij}\ln(1-\delta+\delta\exp(\sigma\sqrt{2}u^{(k)}))+
q_{ij}\ln(1-\delta+\delta\exp(\sigma u^{(k)}-\sigma^2/2))+
q_{ij}\ln(1-\delta+\delta\exp(-\sigma u^{(k)}+\sigma^2/2))]/\sqrt{M}$.
\item Estimate $\Delta_i^{[EC]}$ as $\widehat{\Delta_i^{[EC]}}=\widehat{\olep_i^{[EC]}}-q_{ij}\widehat{\olep_j^{[EC]}}=
\mean_k[\ln(1-\delta+\delta\exp(\sigma(\tilde{\ddot{b}}_i^{(k)}-\tilde{\ddot{b}}_j^{(k)})+\mu_i-\mu_j))]-
\mean_k[\ln(1-\delta+\delta\exp(\sigma\sqrt{2}u^{(k)}+\mu_i-\mu_j))]+
q_{ij}\mean_k[\ln(1-\delta+\delta\exp(\sigma\sqrt{2}u^{(k)}))],$ which one can straightforwardly show has standard error
$\sqrt{s_1^2+s_2^2}$, where $s_1=\sd_k[\ln(1-\delta+\delta\exp(\sigma(\tilde{\ddot{b}}_i^{(k)}-\tilde{\ddot{b}}_j^{(k)})+\mu_i-\mu_j))]/\sqrt{M}$ 
and $s_2=\sd_k[\ln(1-\delta+\delta\exp(\sigma\sqrt{2}u^{(k)}+\mu_i-\mu_j))-q_{ij}\ln(1-\delta+\delta\exp(\sigma\sqrt{2}u^{(k)}))]/\sqrt{M}$.
\item For $\Delta_i^{[E \nat C]}$, we recall that $\olep_j^{[E \nat C]}=0$, so $\Delta_i^{[E \nat C]}=\olep_i^{[E \nat C]}$. So we use
$\widehat{\olep_i^{[E \nat C]}}$ for $\widehat{\Delta_i^{[E \nat C]}}$.
\item Finally, we can also estimate $\olr_{i \bs i}-q_{ij}\olr_{j \bs i}$ simply as $\widehat{\olr_{i \bs i}}$
since $\olr_{j \bs i}=0$.
As for (\ref{eq:decomp_overall_IGR}) and (\ref{eq:decomp_overall_rjbsi}), it is again the case that (\ref{eq:decomp_overall_delta})
applies for the estimators as well as for the quantities being estimated.

%***DAN: Most but not all of the below is obsolete now, replaced and upgraded by stuff above this point

@@
